---
title: "Homework 11"
author: "Jhanani Ramkumar"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above. For any included plots, make sure they are clearly labeled. You are free to use any plot type that you feel best communicates the results of your analysis.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
``` 

```{r}
options(scipen = 999)
```

## Data
For this homework, we are going to use the [Gapminder](https://www.gapminder.org/data/) data on population (pop), GDP per capita (gdp), and life expectancy (lex). These data were downloaded and compiled using the `gapminder_compile.Rmd` script in the data folder.  

Let's focus our attention on data from the past 100 years, between 1925 and 2025.
```{r message=FALSE, warning=FALSE}
gapminder <- read_csv("data/gapminder.csv") %>% 
  clean_names() %>% 
  filter(between(year, 1925, 2025))
```

1. Use the function(s) of your choice to get an idea of the overall structure of the data, including its dimensions, column names, classes, etc. 
```{r}
glimpse(gapminder)
```

2. How many countries are represented in this dataset?
```{r}
gapminder %>% 
  distinct(country) %>% 
  count()
```

195 different countries are represented in the data.

3. Which five countries have had the largest absolute population growth between 1925-2025? Show this as a table.
```{r}
gapminder %>% 
  filter(year %in% c(1925, 2025)) %>% 
  select(country, year, pop) %>% 
  pivot_wider(names_from = year, 
              values_from = pop, 
              names_prefix = "pop_") %>% 
  mutate(growth = pop_2025 - pop_1925) %>% 
  arrange(desc(growth)) %>% 
  slice_head(n = 5)
```

The five countries with the largest absolute population growth includes India, China, Indonesia, Pakistan, and USA.

4. Make a plot that shows population growth over time for the top country you found in question #3.
```{r}
gapminder %>% 
  filter(country == "India") %>% 
  ggplot(aes(x=year, y=pop))+
  geom_line(color="red")+
  scale_x_continuous(breaks=seq(1925, 2025, by=5), minor_breaks = NULL)+
  theme(axis.text.x = element_text(angle=60, hjust=1))+
  labs(title="Population Growth in India, 1925-2025", 
       x = "Year", 
       y = "Population Growth")
```

Population growth in India has been increasing since 1950.

5. Has global life expectancy changed between 1925 and 2025? Show the min, median, mean, and max for all countries (combined) in the dataset.
```{r}
gapminder %>% 
  filter(year %in% c(1925, 2025)) %>% 
   group_by(year) %>% 
  summarize(
    min_lex = min(lex, na.rm = TRUE),
    median_lex = median(lex, na.rm = TRUE),
    mean_lex = mean(lex, na.rm = TRUE),
    max_lex = max(lex, na.rm = TRUE)
  ) 
```

Overall, the mean, median, minimum, and maximum values for life expectancy have increased greatly by 2025. For example, the average life expectancy went from around 38 years to 74 years. 

6. Make a plot that shows the distribution of life expectancy for all countries in 1925 and 2025 (try using `geom_density`). What do you notice about the shift in life expectancy over time?
```{r}
gapminder %>% 
  filter(year %in% c(1925, 2025)) %>%
  mutate(year=factor(year)) %>% 
  ggplot(aes(x=lex, color=year, fill=year))+
    geom_density(alpha=0.6)+
    labs(title="Distribution of Life Expectancy, 1925 and 2025", 
         x = "Life Expectancy (years)", 
         y = NULL, 
         color="Year", fill="Year")
```

Over time, the life expectancy averages went from around 30-40 years to 70-80 years!

7. We are interested in the relationship between per capita GDP and life expectancy; i.e. does having more money help you live longer in 2025? Show this as a plot. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gapminder %>% 
  filter(year == 2025, country != "Monaco") %>% 
  ggplot(aes(x = gdp, y = lex)) +
  geom_point(alpha = 0.7, color = "blue3") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_log10()+
  labs(title = "GDP per Capita vs Life Expectancy (2025)",
       x = "GDP per Capita (Log Scale)",
       y = "Life Expectancy")
```

There is a strong, positive correlation between GDP per capita and life expectancy. So, yes having more money does increase life expectancy!

8. Which five countries have had the highest absolute GDP per capita growth over the past 50 years? Show this as a table. (Suggestion! Remove Monaco as an extreme outlier.)
```{r}
gdp_top5 <- gapminder %>% 
  filter(year %in% c(1975, 2025), country != "Monaco") %>% 
  select(country, year, gdp) %>% 
  pivot_wider(names_from = year, 
              values_from = gdp, 
              names_prefix = "gdp_") %>% 
  mutate(growth = gdp_2025 - gdp_1975) %>% 
  arrange(desc(growth)) %>% 
  slice_head(n = 5)
gdp_top5
```

The top 5 countries with the highest absolute GDP per capita include Singapore, Ireland, Luxembourg, Guyana, and Norway. 

9. How does per capita GDP growth compare between these same five countries over the past 50 years? Show this as a plot.
```{r}
gapminder %>% 
  filter(country %in% gdp_top5$country,
         year %in% c(1975, 2025)) %>% 
  ggplot(aes(x=year, y=gdp, color = country))+
  geom_line(color="red")+
  geom_point()+
  scale_x_continuous(breaks=seq(1975, 2025, by=5), minor_breaks = NULL)+
  theme(axis.text.x = element_text(angle=60, hjust=1))+
  labs(title="Top 5 Countries With Highest GDP Growth, 1975-2025", 
       x = "Year", 
       y = "GDP Growth per capita")
```

Over the last 50 years, the per capita GDP growth has increased in the same five countries.

10. Do one analysis of your choice that includes a table and plot as outputs.

How does population growth in the last 75 years compare to life expectancy changes?
```{r}
pop_growth <- gapminder %>% 
  filter(year %in% c(1950, 2025)) %>% 
  select(country, year, pop, lex) %>% 
  pivot_wider(names_from = year,
              values_from = c(pop, lex),
              names_sep = "_") %>% 
  mutate(pop_growth = pop_2025 - pop_1950,
         lex_growth = lex_2025 - lex_1950)

pop_growth
```

```{r}
pop_growth %>% 
  ggplot(aes(x = pop_growth, y = lex_growth)) +
  geom_point(alpha = 0.7, color = "blue3") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_log10()+
  labs(title = "Population Growth vs Life Expectancy Change (1950–2025)",
       x = "Population Growth (Log Scale)",
       y = "Life Expectancy Change")
```

There is a somewhat moderate correlation between population growth and life expectancy. There are many other facts that can affect life expectancy which, for example can include, fertility rates and healthcare services. 

## Knit and Upload
Please knit your work as an .html file and upload to Canvas. Homework is due before the start of the next lab. No late work is accepted. Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! 